The role of Similarity in native spoken word recognition

### 2024-07-25 manuscript/manuscript.qmd
New version of introduction

Learning a new language is a non-trivial task, a large part of which is acquiring a new set of word-forms to correspond to existing semantic representations in their lexicon. Imagine that a Spanish listener hears English for the first time and encounters the word /\textipa{"dO:}/ (*door*). Their knowledge of the phonological form in their native language is of little help: /\textipa{"pweR.ta}/ and /\textipa{"dO:}/ do not share any phonemes. It would come as a surprise if the listener was able to translate the English word successfully without having first learnt the word. However, there are some elements in language that may help the learner in acquiring words in a new language. Many languages share some degree of similarity. This may be due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents (also known as cognates) [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographical similarity of the former to the equivalent in their native language. Second language learners have been found to learn cognates more easily than non-cognates [@hall2002automatic; @otwinowska2019more]. This type of cross-linguistic similarity has also been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory].

Event counting on this facilitation effect of form-similarity, listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of potential source of this increased difficulty is the partial mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /\textipa{"pOKt}/ (*porte*), translation of door in French. When comparing the phonemic transcription to the transcription of the Spanish word /\textipa{"pweR.ta}/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /\textipa{K}/ and the open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /\textipa{R}/ and /o/. A perceptual assimilation model has been proposed to account for how speakers may integrate non-native phonemes into their native phonemic categories based on goodness-of-fit [@best2001discrimination; @best1988examination]. Although phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish monolingual may still be able to activate the phonological representation of /\textipa{"pweR.ta}/ (*puerta*) from hearing the word-form /\textipa{"pOKt}/ (*porte*), ultimately allowing them to access the right semantic representation.

@@

Learning a new language is a non-trivial task, a large part of which is acquiring a new set of word-forms to correspond to existing semantic representations in their lexicon. Imagine that a Spanish listener hears English for the first time and encounters the word /dɔː/ (*door*). Their knowledge of the phonological form in their native language is of little help: /pweɾta/ and /dɔː/ do not share any phonemes. It would come as a surprise if the listener was able to translate the English word successfully without having first learnt the word. However, there are some elements in language that may help the learner in acquiring words in a new language. Many languages share some degree of similarity. This may be due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents (also known as cognates) [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographical similarity of the former to the equivalent in their native language. Second language learners have been found to learn cognates more easily than non-cognates [@hall2002automatic; @otwinowska2019more]. This type of cross-linguistic similarity has also been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory].

Event counting on this facilitation effect of form-similarity, listening to non-native speech is costlier than listening to native speech, even for highly proficient bilinguals, and especially in acoustically adverse situations [@lecumberri2010non; @takata1990english]. One of potential source of this increased difficulty is the partial mismatch between the phonology of the native and the non-native languages: some acoustic features embedded in the non-native speech signal do not overlap perfectly with any phonemic category in the listener’s native language. For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /pɔʁt/ (*porte*), translation of door in French. When comparing the phonemic transcription to the transcription of the Spanish word /pweɾta/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /ɾ/ and /o/. A perceptual assimilation model has been proposed to account for how speakers may integrate non-native phonemes into their native phonemic categories based on goodness-of-fit [@best2001discrimination; @best1988examination]. Although phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish monolingual may still be able to activate the phonological representation of /pweɾta/ (*puerta*) from hearing the word-form /pɔʁt/ (*porte*), ultimately allowing them to access the right semantic representation.

@@

While phonological similarity across translation equivalents can help with speech recognition, phonological similarity between words with different meanings can hinder the recognition process. In word retrieval, the presence of phonological neighbours (word-forms that differ from each other in one phoneme) may interfere with the activation and selection of the appropriate lexical representation in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing]. The inhibitory effect of phonological neighbourhoods has also been reported across languages. For instance, @weber2004lexical observed that bilinguals displayed lexical competition effects from words in their native language when performing a word-referent matching task in their L2. The assimilation of non-native phonemes into native categories [@best2001discrimination; @best1988examination] was used to explain how lexical competitors in the native language can be activated even when the target words contain non-native phonemes. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition.

In the present study, we investigate the extent to which monolinguals can exploit phonological similarity between translation equivalents when listening to words in an unfamiliar language, and how it interacts with competition from phonological neighbours in the native language. We used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. We explored listeners’ reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. The effect is predicted to be especially strong if any of the competitors is more lexically frequent than the target translation. We hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech.We conducted a series of three experiments to test these predictions.

@@

>>>>

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (\textipa{/ju: n@U mi:}/) in English, Japanese-native listeners tend to understand *yunomi* (/\textipa{jUnomi}/), which is a type of Japanese teacup. Overall, the occurence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which often results in outrageous mistranslations like the ones aforementioned---homophonic translations sometimes may lead to the activation of the intended meaning (i.e., a correct translation). For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /\textipa{"pOKt}/ (*porte*), translation of door in French. When comparing the phonemic transcription of the French to that of its Spanish translation /\textipa{"pweR.ta}/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /\textipa{K}/ and the open-mid back rounded vowel /\textipa{O}/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /\textipa{R}/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /\textipa{"pweR.ta}/ (*puerta*) from hearing the word-form /\textipa{"pOKt}/ (*porte*), ultimately allowing them to access the correct semantic representation. Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

While Soramimi, and more generally homophonic translation, are popular and frequent phenomena, the psycholinguistic literature on the topic is relatively scarce, and has mostly focused on the phonological variables involved in the occurence of Soramimi. For instance, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi hour. These instances consisted of homophonic translations of English song lyrics to Japanese words and phrases. The vast majority of the reported instances were phrases (96%) and a few, to single words (4%). The phonetic features of the presented English lyrics were preserved with relatively high variability, with some Japanese resulting words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. When analysing the few instances of homophonic translations of single words, the author identified three phonological processes that might explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /\textipa{kraI}/ to *kurai*, /\textipa{kuRai}/ [dark]), deletion (e.g., *go*, /\textipa{goU}/ to *go*, /go/ [go], and alternation (e.g., *low*, /\textipa{loU}/ to *rou*, /\textipa{Roo}/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. Native prosody also constrains the occurence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries *before* stressed syllables (following the more frequent trochaic patter in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries *after* stressed syllables, following the more frequent iambic pattern in French.

While the previous studies have address the phonological aspects of homophonic translation, little attention has been paid to the dynamics of lexical activation and competition that underlay this phenomenon. In the present work we investigated the interplay between phonological similarity and phonological neighbourhood density to examine how listeners of an unfamiliar language activate lexical representations in their native language. In particular, we studied how the presence of phonological neighbours in the native language lexicon affects the extent to which listeners are able to activate the correct translation of an auditorily presented word in an unfamiliar language. We used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. We explored listeners’ reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing]. Since the interference effects of phonological neighbourhood density has also been reported across languages [@weber2004lexical], we hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

@@

When some German speakers listened to the song *The Power* by SNAP! (*Coyote Ugly*, 2000), many of them misheard the line "I've got the power" as *Agatha Bauer!* When Michael Jackson vocalised "ma-ma-coo-sah" in his song *Wanna Be Startin' Somethin'* (*Thriller*, 1982), some Dutch speakers understood *mama appelsap* [mama applejuice]. When Spanish speakers listen to the line "Circumvent your thick ego" from the song *Pictures* by System of a Down (*Steal this Album!*, 2002), they tend to hear *Sácame de aquí, King-Kong* [take me out of here, King-Kong]. When Japanese speakers listen to the line "I want to hold your hand" in the homonym song by The Beatles (*Meet The Beatles*, 1964), they often mishear *Aho na hōnyōhan* [stupid urinator]. Outrageous as these examples might sound to speakers of other languages, many readers may know a few cases in their own native language. This auditory illusion can feel quite real, and often inevitable after the first time it happens. In Japanese, this phenomenon takes the name *Soramimi* (lit. "empty ear"), after the *Soramimi Hour*---a section of the program *Tamori Club* hosted by comedian Morita Kazuyoshi (Tamori)---in which instances of this illusion were presented with comedy purposes. Soramimi is a particular case of *homophonic translation*: words or phrases in one language are translated into similar-sounding words and phrases in a different language, without necessarily preserving the meaning. This phenomenon extends beyond song lyrics, and may occur anytime when listening to non-native speech. For instance, when listening to "You know me?" (/ju: nəʊ mi:/) in English, Japanese-native listeners tend to understand *yunomi* (/jʊnomi/), which is a type of Japanese teacup. Overall, the occurence of Soramimi---and more generally, homophonic translations---suggests that speech in an unfamiliar language has the potential to activate lexical representations in the native language.

Although homophonic translations do not necessarily take place with the preservation of meaning---which often results in outrageous mistranslations like the ones aforementioned---homophonic translations sometimes may lead to the activation of the intended meaning (i.e., a correct translation). For example, imagine a Spanish native with no previous familiarity with any other language listening to French for the first time. This listener may encounter the word /pɔKt/ (*porte*), translation of door in French. When comparing the phonemic transcription of the French to that of its Spanish translation /pweɾta/ (*puerta*), one may be led to infer that these two words do not have much overlap. But one thing to note is that the voiced uvular fricative consonant /ʁ/ and the open-mid back rounded vowel /ɔ/, which do not exist in Spanish as phonemic categories, can be perceived as allophonic variations the native (Spanish) phonemes /ʁ/ and /o/, as described by assimilation models of phoneme perception [@best2001discrimination; @best1988examination]. Although this phonetic mismatch between the non-native word and its translation can have a noticeable toll on comprehension @cutler2004patterns, the Spanish listener may still be able to activate the phonological representation of /pweɾta}/ (*puerta*) from hearing the word-form /pɔʁt/ (*porte*), ultimately allowing them to access the correct semantic representation. Given enough phonological similarity between the non-native word and its translation in the native language, a naïve listener may succeed at word comprehension when listening to an unfamiliar language. Such phonological similarity between translation equivalents---known as *cognateness*---is common across many languages, and is often due to typological closeness and/or socio-historical events involving the speakers of these languages (e.g., migration, social contact). For example, Romance languages such as Spanish and Catalan share many form-similar translation equivalents  [@schepens2012distributions], as in the case of *puerta* and *porta* (*door* in Spanish and Catalan, respectively). Given no prior knowledge of either language, a Spanish native speaker is likely to be much more successful at correctly translating Catalan *porta* than English *door*, due to the phonological and orthographic similarity of the former to the equivalent in their native language.

While Soramimi, and more generally homophonic translation, are popular and frequent phenomena, the psycholinguistic literature on the topic is relatively scarce, and has mostly focused on the phonological variables involved in the occurence of Soramimi. For instance, @otake2007interlingual analysed 194 instances of Soramimi, broadcasted between 1992 and 2007 by the TV show Soramimi hour. These instances consisted of homophonic translations of English song lyrics to Japanese words and phrases. The vast majority of the reported instances were phrases (96%) and a few, to single words (4%). The phonetic features of the presented English lyrics were preserved with relatively high variability, with some Japanese resulting words or phrases sharing high overlap with the original English strings of sounds, and some sharing very little overlap. When analysing the few instances of homophonic translations of single words, the author identified three phonological processes that might explain how Japanese listeners reconstructed the English input to generate the Japanese words: insertion (e.g., *cry* /krʌɪ/ to *kurai*, /kuɾai/ [dark]), deletion (e.g., *go*, /goʊ/ to *go*, /go/ [go], and alternation (e.g., *low*, /loʊ/ to *rou*, /ɾoo/ [wax]). This suggests that the reported homophonic translations can be explained, to some extent, as the result of the Japanese listeners accommodating the strings of English sounds to the Japanese phoneme inventory and Japanese phonotactics [@peperkamp2008perceptual; @dupoux1999epenthetic]. Native prosody also constrains the occurence of homophonic translations. For instance, @kentner2015rhythmic presented in the auditory modality French-native and German-native participants with English songs with lyrics. Homophonic translations followed language-specific segmentation strategies. Most homophonic translations reported by German-native participants resulted from inserting phrase and word boundaries *before* stressed syllables (following the more frequent trochaic patter in German). On the contrary, most homophonic translations reported by French-native participants resulted from inserting phrase and word boundaries *after* stressed syllables, following the more frequent iambic pattern in French.

While the previous studies have address the phonological aspects of homophonic translation, little attention has been paid to the dynamics of lexical activation and competition that underlay this phenomenon. In the present work we investigated the interplay between phonological similarity and phonological neighbourhood density to examine how listeners of an unfamiliar language activate lexical representations in their native language. In particular, we studied how the presence of phonological neighbours in the native language lexicon affects the extent to which listeners are able to activate the correct translation of an auditorily presented word in an unfamiliar language. We used a translation elicitation task in which participants listened to words from an unfamiliar language, and then tried to guess their translation in their native language. We will henceforth refer to the auditory-presented words heard by participants on each trial as presented words, and the correct translation for the presented words as target translations. We explored listeners’ reliance on phonological similarity to succeed in the task by manipulating the amount of phonological similarity between the presented words and their target translations. Since participants were unfamiliar with the presented language, they could only use phonological similarity between the presented language and their native language to successfully translate the words. We therefore predicted that participants’ performance should increase when the translation pairs are phonologically more similar. We also predicted that there is a minimum threshold of phonological similarity to be sufficient for translation.

We further investigated the effect of phonological competitors. Even if the presented word and its target translation share high phonological similarity, participants may still provide incorrect translations if the presented word also shares high phonological similarity with other words in the native language. Words with denser phonological neighbourhoods are recognised more slowly and less accurately than words with sparser phonological neighbourhoods [@dufour2003lexical; @goldinger1989priming; @hamburger1996phonological; @luce1990similarity; @luce1998recognizing]. This is especially true if such neighbours share higher phonological similarity with the presented word, or are lexically more frequent than the target translation [@luce1998recognizing]. Since the interference effects of phonological neighbourhood density has also been reported across languages [@weber2004lexical], we hypothesised that the size of the facilitatory role of phonological similarity would be inversely proportional to the amount of higher-frequency phonological neighbours of the presented word in the target language. In the case of non-native speech recognition, the presence of cross-linguistic pairs which are phonologically similar but differ in meaning (e.g., false friends) may act as distractors during lexical access, obstructing the selection of the appropriate target translation in the listener’s lexicon. For instance, @otwinowska2019more reported that false friends were disadvantaged relative to non-cognates by Polish second language learners, in contrast to cognates which were known better. It is therefore important to investigate the joint effect of both cognates and false friends when investigating the effect cross-linguistic phonological similarity has on word recognition in a foreign language. This could shed light on available strategies and challenges associated with the early stages of second language acquisition. For instance, one could expect English participants to incorrectly translate the Spanish word *botón* as *bottom* instead of as its correct translation *button*, due to combined effect of the close phonological similarity between *botón* and *bottom* along with the high lexical frequency of *bottom* relative to *button*. To test this prediction, we developed a lexical frequency-dependent measure of cross-language phonological neighbourhood density, in which a neighbour is counted only if it is higher frequency and is one phoneme apart from the presented word. If the phonological neighbourhood density of the target translation affects participants’ performance negatively, this would suggest that competitors in the native language affect recognition of non-native words during recognition of foreign speech. We conducted a series of three experiments to test these predictions.

@@

@@

This type of cross-linguistic similarity has also been extremely informative for the study of bilingual lexical processing: words that share high form-similarity with their translations are processed faster and more accurately than words sharing lower phonological similarity. This phenomenon has provided strong evidence in favour of the language-non selective hypothesis of bilingual lexical access, which states that bilinguals activate lexical representations in both languages, even during monolingual situations. This facilitation effect has been reported for comprehension [@dijkstra2010cross; @midgley2011effects; @thierry2007brain], production [@costa2000cognate], learning [@de2000hard; @elias2022crosslanguage; @lotto1998effects; @valente2018does], and translation [@christoffels2006memory].
